Hierarchical recurrent models reveal latent states of neural activity in \textit{C. elegans}

### 2018-01-03 drafts/elife/draft.tex
plotting for o2 figures. adding feature for change in o2

\begin{figure}[t!]
\centering%
\includegraphics[width=6in]{figures/figure1} 
\caption{
  \textit{A probabilistic framework for combining
  information across multiple worms in order to learn a shared
  dynamical system for \celegans.}
  \textbf{a.}~The data consists of observed neural activity collected
  from many different worms.  In each recording we observe a different
  subset of labeled neurons in the head ganglia.
  \textbf{b.}~We model the observed activity as a linear function of
  an underlying, continuous, and typically low-dimensional latent state,
  here shown as a two-dimensional signal for each recording.
  \textbf{c.}~The dynamics of these continuous states are governed by
  a co-evolving discrete state.
  \textbf{d.}~These discrete states index into a library of linear
  dynamics functions for each worm, here shown as maps from~$\reals^2$ to~$\reals^2$.
  Each dynamics function is selectively used in different regions of
  continuous latent space, as shown by the different intensities of
  the vector field.
  \textbf{e.}~To increase robustness to model misspecification, we
  use a multivariate-t distribution for the noise in the continuous
  state updates.
  \textbf{f.}~Each worm has its own dynamics library, but they are
  constrained to be close to the shared, canonical library.  This
  combines information to learn shared structure while also allowing
  for worm-to-worm variability.
  \textbf{g.} Likewise, we allow each worm to have its own observation
  variances to account for individual differences that may arise from,
  for example, variable levels of fluorescent protein expression.
  Missing neurons are denoted by gray boxes. 
}
\label{fig:model}
\end{figure}

@@

\subsection*{Emission model aids in identification and reveals clusters of neurons}

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/figure2} 
\caption{ \textit{The mapping from continuous latent states to
    observations aids in neural identification and reveals clusters of
    neurons.}
  \textbf{a.}~The observed neural activity (solid black lines) in the
  first worm of the \citet{kato2015global} dataset along with the
  smoothed activity given by the model (blue lines). We use
  correlations learned across the population to predict the activity
  of unobserved neurons (those with dotted black lines).
  \textbf{b.}~These predictions are a useful aid for neural identification.
  We withheld the labels of a fraction of neurons and measured the similarity
  between their activity and the models predictions.  The similarity was
  significantly higher for the correct assignments as compared to incorrect
  assignments.
  \textbf{c.}~The probabilistic model offers a principled means of assessing
  the dimensionality of neural activity, even though only a subset of neurons
  are labeled on any given trial.  We find that ten latent dimensions yields
  the highest test likelihood, and that allowing per-worm observation variances
  yields substantial improvements in test performance.
  \textbf{d.}~The per-worm variances show that observation noise can
  vary substantially from one worm to another. Gray squares denote missing
  neurons.
  \textbf{e.}~We can view the linear mapping from latent states to observations
  as either a matrix or as an embedding of neurons.
  \textbf{f.}~Viewed as an embedding, we see that neurons cluster into
  correlated groups with similar tuning to the continuous latent state.
  \textbf{g.}~These groups roughly correspond to neurons at different
  locations along the worm's body.  }
\label{fig:emissions}
\end{figure}

The parameters of the learned model offer novel insights into the
structure of neural activity in the head ganglia that can only be
obtained by pooling information across many partial recordings.  With
this aggregated information, we can smooth noisy observations and
predict the activity of unobserved neurons. \citet{kato2015global}
used a smoothing procedure to estimate temporal principal components.
By contrast, we fit our model directly to the first-order differences
in calcium fluorescence; our only preprocessing step is to correct for
photo-bleaching. Fig~\ref{fig:emissions}a show our results for the first
of the five worms in this dataset. We see that the model smooths out
noise in the observed activity and, moreover, makes predictions about
the activity of neurons that could not be identified in this worm.

The predicted activity is based on the correlations between neurons
learned from many recordings.  These predictions are a useful aid for
identifying unlabeled neurons.  For example, we see that
\textsf{SMBDL} was not found in the first worm
(Fig.~\ref{fig:emissions}a). If the activity of one of the unlabeled
neurons closely matches the model's predictions for \textsf{SMBDL},
then this provides some evidence in favor of assigning it this
label. We tested this idea by withholding the labels of some
confidently-identified neurons and comparing their activity to the
model's predicted activity for all unlabeled neurons.  We measured the
similarity between the actual and predicted activity and found that
correct label, or ``candidate,'' tends to be much more similar than
other incorrect candidates, as summarized in
Fig.~\ref{fig:emissions}b.  Combined with other side-information, like
the location of the neuron, this provides a powerful aid for identifying
neurons.  The more neurons that we can identify, the better we will be
able to learn about the dynamics of neural activity.

Even though each individual recording contains only a subset of neurons,
the hierarchical model can still estimate the dimensionality of
neural activity in the entire head ganglia.  We reserve the last 20\% of
the data (3.6 min) for testing and estimate the dimensionality
of neural activity based on the marginal log likelihood the model assigns
to this held-out data. We find that the test likelihood peaks with a
ten-dimensional latent state (Fig.~\ref{fig:emissions}c).  Moreover, the likelihood is substantially
higher in the model that includes per-worm observation variances
compared to the model with shared variances for all worms. Fig.~\ref{fig:emissions}d
shows the inferred variance for each worm in the \citet{kato2015global}
dataset.  Neurons that were not identified are marked with a gray box.
This shows the extent of the missing data and the need for hierarchical
modeling.

@@

\subsection*{SLDS automatically parse data into discrete states with simple dynamics}

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/figure3} 
\caption{ \textit{The hierarchical SLDS decomposes continuous states
    into segments with simple linear dynamics.}
  \textbf{a.}~Three dimensions of the inferred continuous latent states
  for the first worm, color coded by the corresponding discrete latent
  state.
  \textbf{b.}~We select the number of discrete states and justify the
  recurrent and robust extensions by measuring the marginal likelihood
  of held-out test data. We find that the robust, recurrent model
  with eight states performs best.
  \textbf{c.}~The inferred segmentations align with the manual segmentations
  of~\citet{kato2015global}.
  \textbf{d.}~Moreover, the inferred states are in close correspondence with
  the manually labeled states, which were shown to indicate different behaviors.
  \textbf{e.}~Each discrete state is associated with a canonical set of linear
  dynamics, which can be viewed as mappings from~$\reals^10$ to~$\reals^10$.
  We show the mappings for the first three dimensions here, with vectors placed
  at locations where each state was used.  We see a variety of rotational
  states that can be combined to yield dynamics like those in~\textbf{a.}.
  \textbf{f.}~We simulate these linear dynamics from a variety of starting
  locations for the fourth (green) and highlight four for closer inspection.
  \textbf{g.}~These four trajectories give rise to different activations of
  the neural clusters. We see see a dip then rise in cluster 3, which consists
  of neurons related to ventral turns, consistent with the manual labeling
  of the green state in~\textbf{d}.
}
\label{fig:syllables}
\end{figure}

@@

Beyond simply automating a manual process, the SLDS also provides a
quantitative description of the different states.  In particular, each
state is associated with a set of linear dynamics.
Fig.~\ref{fig:syllables}e shows the canonical dynamics associated with
each of the eight discrete states. The arrows in these vector fields
show where the continuous state is mapped to under the corresponding
discrete state (for the first three latent dimensions).  Arrows are
only drawn at those locations where the discrete state was actually
employed.  We see that these dynamics capture the rotational patterns
of the various loops.

We use these dynamics to simulate how brain activity would evolve
under the different states. Fig.~\ref{fig:syllables}f shows 200
simulations of state 4 from various starting locations.  Four
simulations are highlighted in particular, as denoted by the outlined
circle, triangle, square, and pentagon.  We then map these
trajectories into the space of neural activity and show how they would
activate the five clusters shown in Fig.~\ref{fig:emissions}.  We see
that state 4 corresponds to a pronounced dip followed by a rise in the
activity of cluster 3, alongside a slow ramp in the activity of
clusters 1 and 2. Referring back to Fig~\ref{fig:emissions}a, we see
that cluster 3 consists of \textsf{ALA} and three symmetric pairs of
neurons (\textsf{ASK}, \textsf{RIV}, and \textsf{SMDV}).  The~\textsf{ASK}
pair are sensory neurons responsible for initiating local search behavior,
the \textsf{RIV} pair are polymodal inter- and motor-neurons that specify
the ventral bias of turns, and the \textsf{SMDV} are motor neurons
that initiate ventral turns. As state 4 is in close correspondence with
the ``ventral turn'' state labeled by~\citet{kato2015global}, it makes
sense that these neurons would be activated.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/figure4} 
\caption{ \textit{Discrete state transitions show a non-Markovian,
    chain-like structure.}  \textbf{a.}~Counts of inferred transitions
  from one state to another. The states are permuted to emphasize the
  feed-forward structure.  \textbf{b.}~An alternative view of the
  transitions; each state is a node in a graph and edges denote
  high-probability transitions.  This shows transitions between
  forward and reverse crawling linked by dorsal and ventral turns.
  \textbf{c.}~The standard SLDS assumes Markovian transitions, which
  imply a geometric distribution of state durations (dotted black
  lines), but Fig.~\ref{fig:syllables} clearly demonstrates that state
  transitions occur near the origin.  The recurrent SLDS incorporates
  this dependency and generates state durations (solid black lines)
  that are a much better fit to those inferred from the data (colored
  bars).  }
\label{fig:recurrent}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/figure5} 
\caption{\textit{Incorporating exogenous factors like oxygen level
    into the dynamics model yield differential responses in worms from
    different genetic strains and developmental stages.}
  \textbf{a.}~The observed neurons and their inferred observation variances
  for each of the 41 worms in the~\citet{nichols2017global} dataset.
  \textbf{b.}~Worms of different genetic strains (N2, npr1) and developmental
  stages (pre-lethargus, lethargus) exhibit different levels of overall
  state usage, with blue and yellow much more likely in the lethargus stage.
  \textbf{c.}~As above, the inferred states closely match the manually
  labeled states of~\citet{nichols2017global}, with blue and yellow mapping
  onto the forward (\textsf{Fwd}) and quiescent (\textsf{Q}) states.
  \textbf{d.}~We extended the SLDS to allow exogenous factors like oxygen
  level influence discrete transition probabilities.
  \textbf{e., f.}~Upon fitting the model, we find that oxygen has a limited
  effect on N2 worms and a much stronger effect on npr1 worms. In npr1 worms,
  increasing the oxygen level decreases the probability of quiescent and
  forward states and elicits reversals and turns.
}
\label{fig:o2}
\end{figure}

>>>>

\begin{figure}[t!]
\centering%
\includegraphics[width=6in]{figures/v2/figure1} 
\caption{
  \textit{A probabilistic framework for combining
  information across multiple worms in order to learn a shared
  dynamical system for \celegans.}
  \textbf{a.}~The data consists of observed neural activity collected
  from many different worms.  In each recording we observe a different
  subset of labeled neurons in the head ganglia.
  \textbf{b.}~We model the observed activity as a linear function of
  an underlying, continuous, and typically low-dimensional latent state,
  here shown as a two-dimensional signal for each recording.
  \textbf{c.}~The dynamics of these continuous states are governed by
  a co-evolving discrete state.
  \textbf{d.}~These discrete states index into a library of linear
  dynamics functions for each worm, here shown as maps from~$\reals^2$ to~$\reals^2$.
  Each dynamics function is selectively used in different regions of
  continuous latent space, as shown by the different intensities of
  the vector field.
  \textbf{e.}~To increase robustness to model misspecification, we
  use a multivariate-t distribution for the noise in the continuous
  state updates.
  \textbf{f.}~Each worm has its own dynamics library, but they are
  constrained to be close to the shared, canonical library.  This
  combines information to learn shared structure while also allowing
  for worm-to-worm variability.
  \textbf{g.} Likewise, we allow each worm to have its own observation
  variances to account for individual differences that may arise from,
  for example, variable levels of fluorescent protein expression.
  Missing neurons are denoted by gray boxes. 
}
\label{fig:model}
\end{figure}

@@

\subsection*{Hierarchical state space models smooth noisy recordings and predict missing data}

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v2/figure2part1} 
\caption{ \textit{Hierarchical state space models smooth noisy recordings and
  make predictions about activity of unseen neurons.}
  \textbf{a.}~The observed neural activity for the
  five worms of the \citet{kato2015global} dataset along with the
  smoothed activity under the hierarchical model.  Only a subset of neurons
  are labeled in each worm; the rest are treated as missing data.
  The model uses correlations learned across the population to predict the activity
  of unobserved neurons.
  The neurons have been clustered based on similarity, as shown in Fig.~\ref{fig:clustering}.
  \textbf{b.} The observed activity (green lines) and smoothed activity (black lines)
  for 3 minutes of activity from worm 5, as denoted by the dotted black box in panel a.
  Neurons are in the same order as panels a and c.
  \textbf{c.} Zooming in on 30 seconds of activity denoted by the box in panel b.
}
\label{fig:smoothing}
\end{figure}

The parameters of the learned model offer novel insights into the
structure of neural activity in the head ganglia that can only be
obtained by pooling information across many partial recordings.  With
this aggregated information, we can smooth noisy observations and
predict the activity of unobserved neurons. \citet{kato2015global}
used a smoothing procedure to estimate temporal principal components.
By contrast, we fit our model directly to the first-order differences
in calcium fluorescence; our only preprocessing step is to correct for
photo-bleaching. Fig~\ref{fig:smoothing}a show our results for the 
five worms in this dataset. We see that the model smooths out
noise in the observed activity and, moreover, makes predictions about
the activity of neurons that could not be identified in this worm.

Fig.~\ref{fig:smoothing}b and shows the segment of neural activity
denoted by the dotted black box in a. The neurons are clustered based
on similarity, as discussed in the next section.  Here, the green lines
show the observed activity and the black lines show the smoothed activity.
Neurons that were not identified in this worm do not have observed activity,
but the model still makes predictions about their activity. Fig.~\ref{fig:smoothing}c
zooms in further to show a 30 sec window of activity. This shows the
degree of noise in the raw data.

Even though each individual recording contains only a subset of neurons,
the hierarchical model can still estimate the dimensionality of
neural activity in the entire head ganglia.  We reserve the last 20\% of
the data (3.6 min) for testing and estimate the dimensionality
of neural activity based on the marginal log likelihood the model assigns
to this held-out data. We find that the test likelihood peaks with a
ten-dimensional latent state (Fig.~\ref{fig:supp_emissions}c).  Moreover, the likelihood is substantially
higher in the model that includes per-worm observation variances
compared to the model with shared variances for all worms. Fig.~\ref{fig:supp_emissions}d
shows the inferred variance for each worm in the \citet{kato2015global}
dataset.  Neurons that were not identified are marked with a gray box.
This shows the extent of the missing data and the need for hierarchical
modeling.\todo{make supplement figure}

\subsection*{Predictions aid in neural identification}

\todo[inline]{Make a figure for this result}

\subsection*{Emission model reveals clusters of functionally similar neurons}

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v2/figure2part2} 
\caption{ \textit{The mapping from continuous latent states reveals clusters of
    functionally similar neurons.}
  We can view the linear mapping from latent states to observations
  as either a matrix (\textbf{a.}) or as an embedding of neurons (\textbf{b.}).
  \textbf{c.}~Viewed as an embedding, we see that neurons cluster into
  correlated groups with similar tuning to the continuous latent state.
  \textbf{d.}~The embedding similarity reflects correlations between
  neurons, but since it is learned as part of a hierarchical model, it
  predicts similarities for all pairs of neurons, even those that are
  never simultaneously recorded.  These pairs are denoted by gray boxes
  in the empirical correlation matrix.
  \textbf{e.}~Neurons in the same cluster are not necessarily close
  to one another in space, though we do see that, on average, neurons in cluster 2
  are more anterior whereas neurons in clusters 5 and 6 are more posterior.}
\label{fig:clustering}
\end{figure}

% --------- neural identification ----------------
% The predicted activity is based on the correlations between neurons
% learned from many recordings.  These predictions are a useful aid for
% identifying unlabeled neurons.  For example, we see that
% \textsf{SMBDL} was not found in the first worm
% (Fig.~\ref{fig:emissions}a). If the activity of one of the unlabeled
% neurons closely matches the model's predictions for \textsf{SMBDL},
% then this provides some evidence in favor of assigning it this
% label. We tested this idea by withholding the labels of some
% confidently-identified neurons and comparing their activity to the
% model's predicted activity for all unlabeled neurons.  We measured the
% similarity between the actual and predicted activity and found that
% correct label, or ``candidate,'' tends to be much more similar than
% other incorrect candidates, as summarized in
% Fig.~\ref{fig:emissions}b.  Combined with other side-information, like
% the location of the neuron, this provides a powerful aid for identifying
% neurons.  The more neurons that we can identify, the better we will be
% able to learn about the dynamics of neural activity.

@@

\subsection*{SLDS automatically parse data into discrete, behaviorally meaningful states}

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v2/figure3part1} 
\caption{ \textit{The hierarchical SLDS automatically segments neural
  activity based on its dynamics.}
  \textbf{a.}~Three dimensions of the inferred continuous latent states
  for each of the five worms, color coded by the corresponding discrete latent
  state.
  \textbf{b.}~We select the number of discrete states and justify the
  hierarchcial, robust, and recurrent extensions by measuring the marginal likelihood
  of held-out test data. We find that the hierarchical, robust, recurrent model
  with~$K=8$ states performs best.
  \textbf{c.}~The inferred segmentations align with the manual segmentations
  of~\citet{kato2015global}.
  \textbf{d.}~Moreover, the inferred states are in close correspondence with
  the manually labeled states, which were shown to indicate different behaviors.
}
\label{fig:syllables}
\end{figure}

@@

\subsection*{Quantifying the dynamics of neural activity in each discrete state}
\begin{figure}[h!]
\centering
\includegraphics[width=6in]{figures/v2/figure3part2} 
\caption{ \textit{Each state is characterized by simple linear dynamics.}
    \textbf{a.}~Each discrete state is associated with a canonical set of linear
  dynamics, which can be viewed as mappings from~$\reals^{10}$ to~$\reals^{10}$.
  We show the mappings for the first three dimensions here, with vectors placed
  at locations where each state was used.
  \textbf{b.}~We simulate these linear dynamics from a variety of starting
  locations and compute the time course of resulting activations for each
  cluster of neurons.
  \textbf{c.}~The eigenspectra of the dynamics matrices highlight the rotational
  vs. contractive nature of the states and illustrate worm-to-worm variability. 
}
\label{fig:dynamics}
\end{figure}

Beyond simply automating a manual process, the SLDS also provides a
quantitative description of the different states.  In particular, each
state is associated with a set of linear dynamics.
Fig.~\ref{fig:dynamics}a shows the canonical dynamics associated with
each of the eight discrete states. The arrows in these vector fields
show where the continuous state is mapped to under the corresponding
discrete state (for the first three latent dimensions).  Arrows are
only drawn at those locations where the discrete state was actually
employed.  We see that these dynamics capture the rotational patterns
of the various loops.

We use these dynamics to simulate how brain activity would evolve
under the different states. Fig.~\ref{fig:dynamics}b shows how four
simulations of each state from various starting locations  map 
into activations in the space of neural activity. In particular, we show how they would
activate the six clusters shown in Fig.~\ref{fig:clustering}.  We see
that state 4 corresponds to a pronounced dip followed by a rise in the
activity of cluster 3, alongside a slow ramp in the activity of
clusters 1 and 2.
% Referring back to Fig~\ref{fig:emissions}a, we see
% that cluster 3 consists of \textsf{ALA} and three symmetric pairs of
% neurons (\textsf{ASK}, \textsf{RIV}, and \textsf{SMDV}).  The~\textsf{ASK}
% pair are sensory neurons responsible for initiating local search behavior,
% the \textsf{RIV} pair are polymodal inter- and motor-neurons that specify
% the ventral bias of turns, and the \textsf{SMDV} are motor neurons
% that initiate ventral turns. As state 4 is in close correspondence with
% the ``ventral turn'' state labeled by~\citet{kato2015global}, it makes
% sense that these neurons would be activated.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v2/figure4v2} 
\caption{ \textit{Discrete state transitions show a non-Markovian,
    chain-like structure.}  \textbf{a.}~Counts of inferred transitions
  from one state to another. The states are permuted to emphasize the
  feed-forward structure.  \textbf{b.}~An alternative view of the
  transitions; each state is a node in a graph and edges denote
  high-probability transitions.  This shows transitions between
  forward and reverse crawling linked by dorsal and ventral turns.
  \textbf{c.}~The standard SLDS assumes Markovian transitions, which
  imply a geometric distribution of state durations (dotted black
  lines), but Fig.~\ref{fig:syllables} clearly demonstrates that state
  transitions occur near the origin.  With the exception of state 7,
  which is unstable, the recurrent SLDS incorporates
  this dependency and generates state durations (solid black lines)
  that are a better fit to those inferred from the data (colored
  bars).  }
\label{fig:recurrent}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/v1/figure5} 
\caption{\textit{Incorporating exogenous factors like oxygen level
    into the dynamics model yield differential responses in worms from
    different genetic strains and developmental stages.}
  \textbf{a.}~The observed neurons and their inferred observation variances
  for each of the 41 worms in the~\citet{nichols2017global} dataset.
  \textbf{b.}~Worms of different genetic strains (N2, npr1) and developmental
  stages (pre-lethargus, lethargus) exhibit different levels of overall
  state usage, with blue and yellow much more likely in the lethargus stage.
  \textbf{c.}~As above, the inferred states closely match the manually
  labeled states of~\citet{nichols2017global}, with blue and yellow mapping
  onto the forward (\textsf{Fwd}) and quiescent (\textsf{Q}) states.
  \textbf{d.}~We extended the SLDS to allow exogenous factors like oxygen
  level influence discrete transition probabilities.
  \textbf{e., f.}~Upon fitting the model, we find that oxygen has a limited
  effect on N2 worms and a much stronger effect on npr1 worms. In npr1 worms,
  increasing the oxygen level decreases the probability of quiescent and
  forward states and elicits reversals and turns.
}
\label{fig:o2}
\end{figure}
